### 2015-02-11 doc/thesis/system.tex
Improving the text about typing rules

Filter types are subtypes only of themselves and of $\Value$.
More precisey, a filter type $\phi(t_{1},t_{2})$ is a subtype of
the same filter type $\phi(t_{1},t_{2})$, which shares the same
types $t_{1}$ and $t_{2}$, and it is also a subtype of $\Value$.

@@

We also include two kinds of type coercions in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

@@

\subsection{Assignment and function application}

@@

\subsection{Tables and refinement}

The simples expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env, \penv \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type produces a
\emph{closed} table type, as defined by rule \textsc{T-IDREAD}:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env(n) = t}
      {\env, \penv \vdash n_{e}:close(t), \env[n \mapsto open(t)]}
\end{array}
\]

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as defined by
rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env(n) = t}
      {\env, \penv \vdash n_{l}:close(t), \env[n \mapsto close(t)]}
\end{array}
\]

@@

We also have different rules for type checking table indexing to avoid
changing table types in these operations:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
       \env_{1}, \penv \vdash e_{2}:t, \env_{2} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1}, \penv \vdash n[e_{2}]:v_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\end{array}
\]

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{regular} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a coercion expression that allows us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-COERCION} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-COERCION}\\
\dfrac{\env_{1}(n) \subtype t \;\;\;
       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
\end{array}
\]

The rule \textsc{T-COERCION} allows our type system to type check the following example:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{regular} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

We can use $a$ to initialize $b$ because the coercion expression converts
the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$.
We can continue to refine the type of $a$ after the aliasing,
as it still holds an \emph{open} table.
At the end of this example, $a$ has type
$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check another scope.
The rule \textsc{T-FUNCTION3} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION3}\\
\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \rightarrow r, closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))}
\end{array}
\]

The rule \textsc{T-FUNCTION3} uses the predicates \emph{closeall}
and \emph{closeset} to close \emph{open} and \emph{unique} table types.
The function \emph{closeall} closes all \emph{open} and \emph{unique}
table types before type checking another scope.
The function \emph{closeset} closes a given set of free assigned variables,
which is given by the function \emph{fav}.
This rule prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1; \; \mathbf{return} \; x + y$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``z"] \; {<}\String{>} = ``foo"$}
\end{tabular}
\end{center}

\subsection{Objects}

Typed Lua also uses the refinement of table types to type check
common idioms that Lua programmers use for building modules and objects.
For instance, the rule \textsc{T-METHOD3} illustrates how our type system
type checks method definitions:
\[
\begin{array}{c}
\mylabel{T-METHOD3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
       \env_{1}, \penv \vdash n_{2} : l \;\;\;
       \not \exists i \in 1..n \; l \lesssim k_{i}\\
       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}\\
       t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
       \end{array}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
       closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

The rule \textsc{T-METHOD3} allows our type system to type check the
following example, which is a translation of the definition of the class
\texttt{Shape} from Section \ref{sec:oop}:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; o:\Self = setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; o[``x"] = x; \; o[``y"] = y; \; \mathbf{return} \; o$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y;$}\\
& & \multicolumn{2}{l}{$\mathbf{return} \; \mathbf{nothing}$}
\end{tabular}
\end{center}

This example uses the type $\{``x":\Number, ``y":\Number\}_{open}$ to
initialize the local variable $s$, which represents the type of the class
that we are defining.
Then, it uses two method definitions to include $new$ and $move$ into our class.
At the end of the example, $s$ has type
\begin{align*}
\{ & ``x":\Number, ``y":\Number,\\
   & \Const \; ``new":\Self \times \Number \times \Number \times \Void \rightarrow \Self \times \Void,\\
   & \Const \; ``move":\Self \times \Number \times \Number \times \Void \rightarrow \Void \}_{unique}
\end{align*}

Inside the definition of the method \emph{new} we use \emph{setmetatable}
to initialize the $s$ with type $\Self$.
The rule \textsc{T-SETMETATABLE1} express this idea:
\[
\begin{array}{c}
\mylabel{T-SETMETATABLE1}\\
\dfrac{\env(n) = \Self}
      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\Self, \env}
\end{array}
\]

After we define our class, we can use it to create object instances
of this class and call its methods.
The rule \textsc{T-INVOKE1} handles the case where method calls
are expressions that return multiple values:
\[
\begin{array}{c}
\mylabel{T-INVOKE1}\\
\dfrac{\env_{1}, \penv \vdash e:t, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
       \env_{3}, \penv \vdash el:p_{2}, \env_{4} \;\;\;
       \Self \times p_{2} \lesssim p_{1}}
      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
\end{array}
\]

We also use the rule \textsc{T-INVOKE1} as the base case for the rules
that handle the cases where method calls are either statements
or expressions that return only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMINVOKE1}\\
\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e{:}n(el)_{s},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPINVOKE1}\\
\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e{:}n(el)_{e}:first(r), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMINVOKE1} discards the returned values,
while the rule \textsc{T-EXPINVOKE1} uses the predicate \emph{first} to
ensure that only one value is returned.

As an example, assuming that the object $o$ is in the environment and
has the type of the class from the previous example, the method call
\[
o{:}move(10, 10)_{s}
\]
type checks through the rules \textsc{T-STMINVOKE1} and \textsc{T-INVOKE1},
because
\[
\Self \times 10 \times 10 \times \Nil{*} \lesssim \Self \times \Number \times \Number \times \Void
\]

\subsection{Filters and projections}

Typed Lua includes five typing rules for handling the $\mathbf{or}$
logical operator and its common idioms:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-OR1}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t, \env_{2} \;\;\;
       \Nil \not\lesssim t \;\;\;
       \False \not\lesssim t}
      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-OR2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-OR3}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:\False, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-OR4}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:\Nil \cup \False, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t, \env_{3}}
      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-OR5}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
      {\env_{1}, \penv \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-OR1} is the rule that defines the short circuit.
We use the consistent-subtyping relation in this rule to guarantee that
the type system checks the second expression when the first one
has the dynamic type, as it can be hiding a false value.

The rules \textsc{T-OR2}, \textsc{T-OR3}, and \textsc{T-OR4} guarantee
that the final result is the type of the second expression, because the
first one is certainly a false value.

The rule \textsc{T-OR5} is the most general rule, but it is also
the rule that handles the common $\mathbf{or}$ idioms.
It uses the predicate \emph{filter} to filter possible false values
that might be part of the type of the first expression.
We can use pattern matching to the define the recursive predicate
\emph{filter} as follows:
\begin{align*}
filter(t_{1} \cup t_{2}, t_{1}) & = filter(t_{2}, t_{1})\\
filter(t_{1} \cup t_{2}, t_{2}) & = filter(t_{1}, t_{2})\\
filter(t_{1} \cup t_{2}, t_{3}) & = filter(t_{1}, t_{3}) \cup filter(t_{2}, t_{3})\\
filter(t_{1}, t_{2}) & = t_{1}
\end{align*}

The rule \textsc{T-OR5} allows our type system to type check the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
\end{tabular}
\end{center}

Without the \emph{filter} predicate,
the expression $x \; \mathbf{or} \; ``Hello"$ would have type
$\String \cup \Nil \cup ``Hello"$.
The \emph{filter} predicate removes the type $\Nil$ from the result,
leaving the type $\String \cup ``Hello"$.
At the end of the type checking, the expression $x \; \mathbf{or} \; ``Hello"$
has type $\String$ because union types are disjoint and $``Hello" \subtype \String$.

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
The rule \textsc{T-IF2} shows the case where our type system
discriminates the type of a local variable base on the tag \texttt{string}:
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = t\\
       closeall(\env_{1}[n \mapsto \phi(t,\String)]), \penv \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}[n \mapsto \phi(t,filter(t, \String))), \penv \vdash s_{2}, \env_{3}\\
       \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}

>>>>

Filter types are subtypes only of themselves and of $\Value$.
More precisely, a filter type $\phi(t_{1},t_{2})$ is a subtype of
the same filter type $\phi(t_{1},t_{2})$, which shares the same
types $t_{1}$ and $t_{2}$, and it is also a subtype of $\Value$.

@@

We also include two kinds of type coercions in our core language:
the left-hand value $n[c] \; {<}t{>}$ and the expression ${<}t{>} \;n$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

@@

\subsection{Assignment and function application}
\label{sec:assignment}

@@

\subsection{Tables and refinement}
\label{sec:refinement}

The simplest expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env, \penv \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type produces a
\emph{closed} table type, as defined by rule \textsc{T-IDREAD}:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env(n) = t_{1} \;\;\; t_{2} = read(\penv, t_{1})}
      {\env, \penv \vdash n_{e}:close(t_{2}), \env[n \mapsto open(t_{1})]}
\end{array}
\]

In Section \ref{sec:fap} we will explain that \textsc{T-IDREAD} uses
the auxiliary function \emph{read} because it may be accessing an
identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, our type system includes
a small set of \texttt{type} predicates that allow programmers to
discriminate union types.
In Section \ref{sec:fap} we will also explain that our type system
uses filter and projection types in the definition of these predicates
to handle the discrimination of unions of first-level and second-level
types.

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as defined by
rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env(n) = t_{1} \;\;\; t_{2} = write(t_{1})}
      {\env, \penv \vdash n_{l}:close(t_{2}), \env[n \mapsto close(t_{2})]}
\end{array}
\]

In Section \ref{sec:fap} we will explain that \textsc{T-IDWRITE} uses
the auxiliary function \emph{write} because it may be accessing an
identifier that is bound to a filter or projection type.
As we mentioned in Section \ref{sec:unions}, assignments restore
discriminated union types to their original types, and
function \emph{write} also works in this purpose.

@@

We also have different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = t \;\;\;
       read(\penv, t) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:k, \env_{2} \;\;\;
       \exists i \in 1{..}n \; k \lesssim k_{i}
       \end{array}}
      {\env_{1}, \penv \vdash n[e_{2}]:v_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:k, \env_{3} \;\;\;
       \exists i \in 1{..}n \; k \lesssim k_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\end{array}
\]

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{closed} or \emph{regular} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression that
obeys the \textsc{T-COERCE} rule below:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(n) \subtype t \;\;\;
       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
\end{array}
\]

The rule \textsc{T-COERCE} allows our type system to type check the following example:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{regular} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

We can use $a$ to initialize $b$ because the coercion converts
the type of $a$ from $\{``x":\String, ``y":\String\}_{unique}$ to
$\{``x":\String, ``y":\String \cup \Nil\}_{open}$, and results in
$\{``x":\String, ``y":\String \cup \Nil\}_{closed}$,
which is a subtype of
$\{``x":\String, ``y":\String \cup \Nil\}_{regular}$, the type of $b$.
We can continue to refine the type of $a$ after aliasing it to $b$,
as it still holds an \emph{open} table.
At the end of this example, $a$ has type
$\{``x":\String, ``y":\String \cup \Nil, ``z":\Integer\}_{open}$.

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
The rule \textsc{T-FUNCTION3} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION3}\\
\dfrac{closeall(\env_{1}[\vec{n} \mapsto \vec{t}]), \penv[\ret \mapsto r] \vdash s, \env_{2}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}t}){:}r \; s:\vec{t} \times \Void \rightarrow r,\\
       openset(closeset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

Rule \textsc{T-FUNCTION3} uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{closeall} closes all \emph{unique} and \emph{open}
table types.
The function \emph{closeset} closes a given set of free assigned variables,
which is given by the function \emph{fav}.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

This rule also extends the environment $\penv$, bounding the special
variable $\ret$ to the return type $r$.
Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
$\penv$ to type check return statements:
\[
\begin{array}{c}
\mylabel{T-RETURN}\\
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \penv(\ret) = r_{2} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
\end{array}
\]

As an example, rule \textsc{T-FUNCTION3} prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])_{s}$}
\end{tabular}
\end{center}

This example does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at this point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{closeall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{closeset}.
Then, the function \emph{rv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

\subsection{Classes and objects}
\label{sec:cao}

Typed Lua also relies on the refinement of table types to type check
the idioms that Lua programmers use for incrementally building
classes and objects.
For instance, rule \textsc{T-METHOD3} illustrates how our type system
type checks method definitions:
\[
\begin{array}{c}
\mylabel{T-METHOD3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n_{1}) = t_{s} \;\;\; t_{s} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open}\\
       \env_{1}, \penv \vdash n_{2} : l, \env_{2} \;\;\;
       \not \exists i \in 1..n \; l \lesssim k_{i}\\
       closeall(\env_{1}[self \mapsto \Self, \vec{n} \mapsto \vec{t}, \self \mapsto t_{s}]),
       \penv[\ret \mapsto r] \vdash s, \env_{3}\\
       t_{o} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \times \Void \rightarrow r\}_{unique|open}
       \end{array}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
       openset(closeset(\env_{1}[n_{1} \mapsto t_{o}], fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s)),rv(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

Rule \textsc{T-METHOD3} allows our type system to type check the
following example, which is a translation of the definition of the class
\texttt{Shape} from Section \ref{sec:oop}:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; Shape{:}\{``x"{:}\Number, ``y"{:}\Number\}_{unique} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; o:\Self = setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; o[``x"] = x; \; o[``y"] = y; \; \mathbf{return} \; o$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; Shape{:}move (x:\Number, y:\Number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y;$}\\
& & \multicolumn{2}{l}{$\mathbf{return} \; \mathbf{nothing}$}
\end{tabular}
\end{center}

This example uses the type $\{``x":\Number, ``y":\Number\}_{unique}$ to
initialize the local variable $Shape$, which represents the class
that we are defining.
Then, it uses two method definitions to include $new$ and $move$ in our class.
These definitions refine the type of $Shape$, and it becomes the type below:
\begin{align*}
\{ & ``x":\Number, ``y":\Number,\\
   & \Const \; ``new":\Self \times \Number \times \Number \times \Void \rightarrow \Self \times \Void,\\
   & \Const \; ``move":\Self \times \Number \times \Number \times \Void \rightarrow \Void \}_{unique}
\end{align*}

Inside the definition of the methods \emph{new} and \emph{move},
we index variables that have type $\Self$.
This is possible because we include the rule \textsc{T-SELF}:
\[
\begin{array}{c}
\mylabel{T-SELF}\\
\dfrac{\env_{1}, \penv \vdash e:\Self, \env_{2} \;\;\;
       \env_{2}(\self) = t}
      {\env_{1}, \penv \vdash e:t, \env_{2}}
\end{array}
\]

When an expression has type $\Self$, our type system uses this
rule to produce the type that is bound to $\self$ in $\env$.
The variable $\self$ is a special variable that method definitions
and method calls use to bind the type of $\Self$ in the environment.

Inside the definition of the method \emph{new}, we use \emph{setmetatable}
to initialize $o$ with type $\Self$.
The rule \textsc{T-SETMETATABLE1} express this idea:
\[
\begin{array}{c}
\mylabel{T-SETMETATABLE1}\\
\dfrac{\env(n) = \Self}
      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\Self, \env}
\end{array}
\]

In Section \ref{sec:oop} we also mentioned that our type system
uses \emph{setmetatable} to handle single inheritance.
Rule \textsc{T-SETMETATABLE2} allows refining the type of a new
class based on the type of an already defined class,
while rule \textsc{T-SETMETATABLE3} allows overriding a constructor:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-SETMETATABLE2}\\
\dfrac{\env(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}}
      {\env, \penv \vdash setmetatable(\{\}, \{[``\string_\string_index"] = n\}):\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-SETMETATABLE3}\\
\dfrac{\env_{1}, \penv \vdash e = t, \env_{2} \;\;\;
       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \;\;\;
       \env_{1}(n) = \Self \;\;\; \env_{1}(\sigma) \subtype t}
      {\env_{1}, \penv \vdash setmetatable(e, \{[``\string_\string_index"] = n\}):\Self, \env_{2}[\sigma \mapsto t]}
\end{array}
\end{array}
\]

After we define our class, we can use it to create instances
of this class and call their methods.
Rule \textsc{T-INVOKE1} handles the case where method calls
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-INVOKE1}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e:t_{s}, \env_{2}\\
       \env_{2}[\sigma \mapsto t_{s}], \penv \vdash e[n]:\Const \; r_{1} \rightarrow r_{2}, \env_{3}\\
       \env_{3}[\sigma \mapsto t_{s}], \penv \vdash el:r_{3}, \env_{4} \;\;\;
       t_{s} \times r_{3} \lesssim r_{1}
       \end{array}}
      {\env_{1}, \penv \vdash e{:}n(el)_{m}:[\sigma \mapsto t]r_{2}, \env_{4}}
\end{array}
\]

We also use the rule \textsc{T-INVOKE1} as the base case for the rules
that handle the cases where method calls are either statements
or expressions that produce only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMINVOKE1}\\
\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e{:}n(el)_{s},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPINVOKE1}\\
\dfrac{\env_{1}, \penv \vdash e{:}n(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e{:}n(el)_{e}:first(r), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMINVOKE1} discards the produced values,
while the rule \textsc{T-EXPINVOKE1} uses the predicate \emph{first} to
ensure that only one value is produced.

As an example, assuming that the object $o$ is in the environment and
has the type of the class $Shape$, the method call
\[
o{:}move(10, 10)_{m}
\]
type checks through the rule \textsc{T-INVOKE1}, because
\[
Shape \times 10 \times 10 \times \Nil{*} \lesssim Shape \times \Number \times \Number \times \Void
\]
assuming that $Shape$ is an alias to the type of the class $Shape$.
Note that rule \textsc{T-INVOKE1} uses the object type to bind
the type $\Self$ before type checking a method call.

\subsection{Filters and projections}
\label{sec:fap}

In Section \ref{sec:unions} we mentioned that our type system 
includes union types to encode three common Lua idioms:
the use of optional values, the overloading based on the tags of
input parameters, and the overloading on the return type of functions.
Now, we will discuss how our type system uses the $\mathbf{or}$
operator to handle optional values, how it uses filter types to
discriminate unions of first-level types, and how it uses
projection types to discriminate overloaded return types,
which are unions of second-level types.
